---
title: "Visualizing phylogenetic data with ggtree"
output: learnr::tutorial
runtime: shiny_prerendered
---


```{r setup, include=FALSE}
library(learnr)
library(treeio)
library(ggtree)
```



## Visualizing a tree

```{r rtree, exercise = TRUE}
library(ggtree)
set.seed(2024-07-18)
t1 <- rtree(30)

t1
```

```{r tree-layout, exercise = TRUE, exercise.setup="rtree", fig.width=4, fig.height=5}
# try different layouts:
#   rectangular, dendrogram, slanted, ellipse, roundrect
#   fan, circular, inward_circular, radial, equal_angle
#   daylight, ape, igraph::layout_xxx

ggtree(t1) 
# ggtree(t1, layout = 'roundrect') 
# ggtree(t1, layout = 'ellipse') 
# ggtree(t1, layout = 'inward_circular') + xlim_tree(10)
# ggtree(t1, layout = igraph::layout_with_kk) + geom_tiplab()
```



## A layered grammar of graphics


```{r nhx-file, exercise=TRUE}
library(treeio)

tree_file <- system.file("extdata/NHX/ADH.nhx", package='treeio')
tree_file
```

```{r nhx-tree, exercise=TRUE, exercise.setup="nhx-file"}
tree <- read.nhx(tree_file)
tree
```

### 1. tree structure

```{r vis-nhx, exercise=TRUE, exercise.setup="nhx-tree"}
library(ggtree)
p <- ggtree(tree)
p
```

### 2. tip labels

```{r vis-nhx2, exercise=TRUE, exercise.setup="vis-nhx"}
p <- p + geom_tiplab()
p
```

### 3. displays the species

```{r vis-nhx3, exercise=TRUE, exercise.setup="vis-nhx2", message=FALSE, warning=FALSE}
p <- p + geom_label(aes(x=branch, label=S), fill='lightgreen')
p
```

### 4. displays duplication events

```{r vis-nhx4, exercise=TRUE, exercise.setup="vis-nhx3", message=FALSE, warning=FALSE}
p <- p + geom_label(aes(label=D), fill='firebrick')
p
```


### 4. displays bootstrap values

```{r vis-nhx5, exercise=TRUE, exercise.setup="vis-nhx4", message=FALSE, warning=FALSE}
p <- p + geom_text(aes(label=B), hjust = -.5)
p
```

## ggtree examples

### Visualizing bootstrap values as symbolic points

### 1. Visualizing the tree

```{r bp, exercise=TRUE, fig.width=8, fig.height=12}
library(ggtree)
library(treeio)
library(TDbook)
library(ggplot2)

tree <- read.newick(text = text_RMI_tree, node.label = 'support')

p <- ggtree(
  tree,
  size = 1.5, 
  right = TRUE 
)

p1 <- p + 
  geom_tiplab(
    size = 4.5,  
    hjust = -0.060, 
    fontface = "bold" 
    ) +
  xlim(0, .09) 

p1
```

### Visualizing bootstrap values as a continuous variable


```{r bp2, exercise=TRUE, exercise.setup="bp", fig.width=8, fig.height=12}
p2 <- p1 +
  geom_nodepoint(
    data = td_filter(parent != node), 
    mapping = aes(fill = support), 
    size = 3.5, 
    shape = 21 
    )

p2
```

### Visualizing bootstrap values as a discrete variable

```{r bp3, exercise=TRUE, exercise.setup="bp", fig.width=8, fig.height=12}
p2 <- p1 +
  geom_nodepoint(
    data = td_filter(parent != node),
    mapping = aes(fill = cut(support, c(0, 700, 900, 1000))), 
    size = 3.5,
    shape = 21
    )
p2
```

### Adjusting details

```{r bp4, exercise=TRUE, exercise.setup="bp3", fig.width=8, fig.height=12, message=FALSE, warning=FALSE}
p3 <- p2 + scale_fill_manual( # 更改点的颜色
  values = c("white", "grey", "black"), # 点的颜色
  guide = 'legend', 
  name = 'Bootstrap Percentage(BP)', # legend标题
  breaks = c('(900,1e+03]', '(700,900]', '(0,700]'), 
  labels = expression(BP>=90,70 <= BP * " < 90", BP < 70) # 调整每个颜色的代表名字
  )

p4 <- p3 + theme(legend.position=c(0.2, 0.2))
p5 <- p4 + geom_treescale()

p5
```

## Annotating tree with data

### 1. Estimating ancestral states and visualizing state transition

```{r anc, exercise = TRUE, fig.width=15, fig.height=10, message=FALSE, warning=FALSE, out.width="100%"}
library(ggimage)
library(ggtree)
library(TDbook)
library(ggnewscale)

# load `tree_boots`, `df_tip_data`, and `df_inode_data` from 'TDbook'

mass <- setNames(df_tip_data$mass_in_kg, df_tip_data$Newick_label)
fit <- phytools::fastAnc(tree_boots, mass, vars=TRUE, CI=TRUE)

td <- data.frame(node = nodeid(tree_boots, names(mass)), trait = mass)
nd <- data.frame(node = as.numeric(names(fit$ace)), trait = fit$ace)
d <- rbind(td, nd)
tree <- full_join(tree_boots, d, by='node')

p <- ggtree(tree, aes(color=trait), continuous='color', size=3) +
  scale_color_gradientn(colours=c("red", "orange", "green", "cyan", "blue"), 
                        name="mass (kg)") + 
  geom_tiplab(hjust = -.2) + xlim(-.1, 4)

p 
```

### 2. Visualizing tree with taxa information
```{r anc2, exercise = TRUE, exercise.setup='anc', fig.width=15, fig.height=10, message=FALSE, warning=FALSE, out.width="100%"}
p2 <- p %<+% df_tip_data + 
    new_scale_color() +
    geom_tippoint(aes(shape = trophic_habit, color = trophic_habit), size=8) + 
    theme(legend.position = "right") + 
    scale_size_continuous(range = c(3, 10))

p2
```

### 3. Visualizing tree with internal node information

```{r anc3, exercise = TRUE, exercise.setup='anc2', fig.width=15, fig.height=10, message=FALSE, warning=FALSE, out.width="100%"}
p3 <- p2 %<+% df_inode_data + 
    geom_label(aes(label = vernacularName.y, fill = posterior)) + 
    scale_fill_gradientn(colors = RColorBrewer::brewer.pal(3, "YlGnBu"))

p3
```


### 4. Overlay organism silhouettes


```{r anc4, exercise = TRUE, exercise.setup='anc3', fig.width=15, fig.height=10, message=FALSE, warning=FALSE, out.width="100%"}

library(ggimage) 
id <- phylopic_uid(df_tip_data$Newick_label) 

p3 %<+% id + 
  geom_tiplab(aes(image=uid), geom="phylopic", offset=.75, size=.1)

```
## Building a complex plot with ggtree and ggtreeExtra

### 1. Visualize the tree structure
```{r hmptree0, exercise=TRUE}

# load data from TDbook, including 
# tree_hmptree, 
# df_tippoint (the abundance and types of microbes),
# df_ring_heatmap (the abundance of microbes at different body sites),
# and df_barplot_attr (the abundance of microbes of greatest prevalence)

library(TDbook)
tree_hmptree

library(ggtree)
library(ggtreeExtra)
library(ggplot2)
library(ggstar)
library(ggnewscale)

f <- ggtree(tree_hmptree, layout = 'fan') 

f
```

### 2. Mapping data to the tree structure

```{r hmptree, exercise=TRUE, exercise.setup="hmptree0"}
f <- f %<+% df_tippoint + geom_star(
                        mapping=aes(fill=Phylum, starshape=Type, size=Size),
                        position="identity",starstroke=0.1) +
        scale_fill_manual(values=c("#FFC125","#87CEFA","#7B68EE","#808080",
                                "#800080", "#9ACD32","#D15FEE","#FFC0CB",
                                "#EE6A50","#8DEEEE", "#006400","#800000",
                                "#B0171F","#191970"),
                           guide=guide_legend(keywidth = 0.5, 
                                        keyheight = 0.5, order=1,
                                        override.aes=list(starshape=15)),
                           na.translate=FALSE)+
        scale_starshape_manual(values=c(15, 1),
                           guide=guide_legend(keywidth = 0.5, 
                                        keyheight = 0.5, order=2),
                           na.translate=FALSE)+
        scale_size_continuous(range = c(1, 2.5),
                           guide = guide_legend(keywidth = 0.5, 
                                        keyheight = 0.5, order=3,
                                        override.aes=list(starshape=15)))
f
```


### 3.1. Aligning graph to the tree based on tree structure
```{r hmptree-heatmap, exercise=TRUE, exercise.setup="hmptree", message=FALSE, warning=FALSE, fig.width=4, fig.height=8}
head(df_ring_heatmap)
p.heatmap <- ggplot() + 
  geom_tile(data = df_ring_heatmap, 
            mapping = aes(x = Sites, y = ID, fill = Sites, alpha = Abundance)
            ) +
  scale_alpha_continuous(range = c(0, 1)) +
  scale_fill_manual(
    values = c("#0000FF","#FFA500","#FF0000", 
               "#800000", "#006400","#800080","#696969")
    ) +
  theme_tree()

p.heatmap
```

```{r hmptree-heatmap2, exercise=TRUE, exercise.setup="hmptree", message=FALSE, warning=FALSE, fig.width=10, fig.height=8}
f1 <- f + 
  new_scale_fill() +
  geom_fruit(
    data = df_ring_heatmap,
    geom = geom_tile,
    mapping = aes(x = Sites, y = ID, fill = Sites, alpha = Abundance),
    color = "grey50",
    size = 0.02,
    offset = 0.04 # control the spacing between the tree layer and heatmap layer
    ) +
  scale_fill_manual(values=c("#0000FF","#FFA500","#FF0000",
                             "#800000", "#006400","#800080","#696969")
                    ) +
  scale_alpha_continuous(range = c(0, 1)) +
  theme(
    legend.key.height = unit(.5, 'cm'), # legend 大小 
    legend.key.width = unit(.5, 'cm'),
    legend.title = element_text(size = 8),
    legend.text = element_text(size=6)
    ) 

f1
```

### 3.2. Aligning graph to the tree based on tree structure

```{r hmptree-barplot, exercise=TRUE, exercise.setup="hmptree", message=FALSE, warning=FALSE, fig.width=4, fig.height=8}
p.barplot <- ggplot() + 
  geom_col(
    data = df_barplot_attr, 
    mapping = aes(x=HigherAbundance, y = ID, fill=Sites)
    ) + 
  scale_fill_manual(values=c("#0000FF","#FFA500","#FF0000", 
                             "#800000", "#006400","#800080","#696969")
                    ) + 
  theme_tree2()

p.barplot
```

```{r hmptree-barplot2, exercise=TRUE, exercise.setup="hmptree-heatmap2", fig.width=10, fig.height=8, message=FALSE, warning=FALSE}
f2 <- f1 + 
  geom_fruit(
    data = df_barplot_attr,
    geom = geom_col,
    mapping = aes(x=HigherAbundance, y = ID, fill=Sites),
    pwidth = .38, # the space ratio for plotting the barchart relative to the tree layer
    axis.params = list(axis = 'x')
    ) +
  scale_fill_manual(values=c("#0000FF","#FFA500","#FF0000",
                             "#800000", "#006400","#800080","#696969")) 

f2
```

### 4. Transforming circular layout to rectangular layout

```{r hmptree-rect, exercise=TRUE, exercise.setup="hmptree-barplot2", fig.width=10, fig.height=8, message=FALSE, warning=FALSE}
f2 + layout_rectangular()
```
