---
title: "Visualizing phylogenetic data with ggtree"
output: learnr::tutorial
runtime: shiny_prerendered
---


```{r setup, include=FALSE}
library(learnr)
library(treeio)
library(ggtree)
```



## Visualizing a tree

### 1. Generate a random tree with 30 tips
```{r rtree, exercise = TRUE}
library(ggtree)
set.seed(2024-07-18)
t1 <- rtree(30)

t1
```

### 2. Visualize the tree 

```{r tree-layout, exercise = TRUE, exercise.setup="rtree", fig.width=4, fig.height=5}
# try different layouts:
#   rectangular, dendrogram, slanted, ellipse, roundrect
#   fan, circular, inward_circular, radial, equal_angle
#   daylight, ape, igraph::layout_xxx

ggtree(t1) 
# ggtree(t1, layout = 'roundrect') 
# ggtree(t1, layout = 'ellipse') 
# ggtree(t1, layout = 'inward_circular') + xlim_tree(10)
# ggtree(t1, layout = igraph::layout_with_kk) + geom_tiplab()
```

### 3. Export the plot to a file

```{r treesave, exercise = TRUE, exercise.setup="rtree", fig.width=4, fig.height=5}
p <- ggtree(t1, layout = 'ellipse') 
png <- tempfile(fileext = '.png')
png

pdf <- tempfile(fileext = '.pdf')
pdf

ggsave(filename = png, plot = p)
ggsave(filename = pdf, plot = p)
```



## A layered grammar of graphics


```{r nhx-file, exercise=TRUE}
library(treeio)

tree_file <- system.file("extdata/NHX/ADH.nhx", package='treeio')
tree_file
```

```{r nhx-tree, exercise=TRUE, exercise.setup="nhx-file"}
tree <- read.nhx(tree_file)
tree
```

### 1. tree structure

```{r vis-nhx, exercise=TRUE, exercise.setup="nhx-tree"}
library(ggtree)
p <- ggtree(tree)
p
```

### 2. tip labels

```{r vis-nhx2, exercise=TRUE, exercise.setup="vis-nhx"}
p <- p + geom_tiplab()
p
```

### 3. displays the species

```{r vis-nhx3, exercise=TRUE, exercise.setup="vis-nhx2", message=FALSE, warning=FALSE}
p <- p + geom_label(aes(x=branch, label=S), fill='lightgreen')
p
```

### 4. displays duplication events

```{r vis-nhx4, exercise=TRUE, exercise.setup="vis-nhx3", message=FALSE, warning=FALSE}
p <- p + geom_label(aes(label=D), fill='firebrick')
p
```


### 4. displays bootstrap values

```{r vis-nhx5, exercise=TRUE, exercise.setup="vis-nhx4", message=FALSE, warning=FALSE}
p <- p + geom_text(aes(label=B), hjust = -.5)
p
```



## Visualizing bootstrap values as symbolic points

### 1. Visualizing the tree

```{r bp, exercise=TRUE, fig.width=8, fig.height=12}
library(ggtree)
library(treeio)
library(TDbook)
library(ggplot2)

tree <- read.newick(text = text_RMI_tree, node.label = 'support')

p <- ggtree(
  tree,
  size = 1.5, 
  right = TRUE 
)

p1 <- p + 
  geom_tiplab(
    size = 4.5,  
    hjust = -0.060, 
    fontface = "bold" 
    ) +
  xlim(0, .09) 

p1
```

### Visualizing bootstrap values as a continuous variable


```{r bp2, exercise=TRUE, exercise.setup="bp", fig.width=8, fig.height=12}
p2 <- p1 +
  geom_nodepoint(
    data = td_filter(parent != node), 
    mapping = aes(fill = support), 
    size = 3.5, 
    shape = 21 
    )

p2
```

### Visualizing bootstrap values as a discrete variable

```{r bp3, exercise=TRUE, exercise.setup="bp", fig.width=8, fig.height=12}
p2 <- p1 +
  geom_nodepoint(
    data = td_filter(parent != node),
    mapping = aes(fill = cut(support, c(0, 700, 900, 1000))), 
    size = 3.5,
    shape = 21
    )
p2
```

### Adjusting details

```{r bp4, exercise=TRUE, exercise.setup="bp3", fig.width=8, fig.height=12, message=FALSE, warning=FALSE}
p3 <- p2 + scale_fill_manual( 
  values = c("white", "grey", "black"), 
  guide = 'legend', 
  name = 'Bootstrap Percentage(BP)', 
  breaks = c('(900,1e+03]', '(700,900]', '(0,700]'), 
  labels = expression(BP>=90,70 <= BP * " < 90", BP < 70) 
  )

p4 <- p3 + theme(legend.position=c(0.2, 0.2))
p5 <- p4 + geom_treescale()

p5
```

## Annotating tree with data

### 1. Estimating ancestral states and visualizing state transition

```{r anc, exercise = TRUE, fig.width=15, fig.height=10, message=FALSE, warning=FALSE, out.width="100%"}
library(ggimage)
library(ggtree)
library(TDbook)
library(ggnewscale)

# load `tree_boots`, `df_tip_data`, and `df_inode_data` from 'TDbook'

mass <- setNames(df_tip_data$mass_in_kg, df_tip_data$Newick_label)
fit <- phytools::fastAnc(tree_boots, mass, vars=TRUE, CI=TRUE)

td <- data.frame(node = nodeid(tree_boots, names(mass)), trait = mass)
nd <- data.frame(node = as.numeric(names(fit$ace)), trait = fit$ace)
d <- rbind(td, nd)
tree <- full_join(tree_boots, d, by='node')

p <- ggtree(tree, aes(color=trait), continuous='color', size=3) +
  scale_color_gradientn(colours=c("red", "orange", "green", "cyan", "blue"), 
                        name="mass (kg)") + 
  geom_tiplab(hjust = -.2) + xlim(-.1, 4)

p 
```

### 2. Visualizing tree with taxa information
```{r anc2, exercise = TRUE, exercise.setup='anc', fig.width=15, fig.height=10, message=FALSE, warning=FALSE, out.width="100%"}
p2 <- p %<+% df_tip_data + 
    new_scale_color() +
    geom_tippoint(aes(shape = trophic_habit, color = trophic_habit), size=8) + 
    theme(legend.position = "right") + 
    scale_size_continuous(range = c(3, 10))

p2
```

### 3. Visualizing tree with internal node information

```{r anc3, exercise = TRUE, exercise.setup='anc2', fig.width=15, fig.height=10, message=FALSE, warning=FALSE, out.width="100%"}
p3 <- p2 %<+% df_inode_data + 
    geom_label(aes(label = vernacularName.y, fill = posterior)) + 
    scale_fill_gradientn(colors = RColorBrewer::brewer.pal(3, "YlGnBu"))

p3
```


### 4. Overlay organism silhouettes


```{r anc4, exercise = TRUE, exercise.setup='anc3', fig.width=15, fig.height=10, message=FALSE, warning=FALSE, out.width="100%"}

library(ggimage) 
id <- phylopic_uid(df_tip_data$Newick_label) 

p3 %<+% id + 
  geom_tiplab(aes(image=uid), geom="phylopic", offset=.75, size=.1)

```



## Building a complex plot with ggtree and ggtreeExtra

### 1. Visualize the tree structure
```{r hmptree0, exercise=TRUE}

# load data from TDbook, including 
# tree_hmptree, 
# df_tippoint (the abundance and types of microbes),
# df_ring_heatmap (the abundance of microbes at different body sites),
# and df_barplot_attr (the abundance of microbes of greatest prevalence)

library(TDbook)
tree_hmptree

library(ggtree)
library(ggtreeExtra)
library(ggplot2)
library(ggstar)
library(ggnewscale)

f <- ggtree(tree_hmptree, layout = 'fan') 

f
```

### 2. Mapping data to the tree structure

```{r hmptree, exercise=TRUE, exercise.setup="hmptree0"}
f <- f %<+% df_tippoint + geom_star(
                        mapping=aes(fill=Phylum, starshape=Type, size=Size),
                        position="identity",starstroke=0.1) +
        scale_fill_manual(values=c("#FFC125","#87CEFA","#7B68EE","#808080",
                                "#800080", "#9ACD32","#D15FEE","#FFC0CB",
                                "#EE6A50","#8DEEEE", "#006400","#800000",
                                "#B0171F","#191970"),
                           guide=guide_legend(keywidth = 0.5, 
                                        keyheight = 0.5, order=1,
                                        override.aes=list(starshape=15)),
                           na.translate=FALSE)+
        scale_starshape_manual(values=c(15, 1),
                           guide=guide_legend(keywidth = 0.5, 
                                        keyheight = 0.5, order=2),
                           na.translate=FALSE)+
        scale_size_continuous(range = c(1, 2.5),
                           guide = guide_legend(keywidth = 0.5, 
                                        keyheight = 0.5, order=3,
                                        override.aes=list(starshape=15)))
f
```


### 3.1. Aligning graph to the tree based on tree structure
```{r hmptree-heatmap, exercise=TRUE, exercise.setup="hmptree", message=FALSE, warning=FALSE, fig.width=4, fig.height=8}
head(df_ring_heatmap)
p.heatmap <- ggplot() + 
  geom_tile(data = df_ring_heatmap, 
            mapping = aes(x = Sites, y = ID, fill = Sites, alpha = Abundance)
            ) +
  scale_alpha_continuous(range = c(0, 1)) +
  scale_fill_manual(
    values = c("#0000FF","#FFA500","#FF0000", 
               "#800000", "#006400","#800080","#696969")
    ) +
  theme_tree()

p.heatmap
```

```{r hmptree-heatmap2, exercise=TRUE, exercise.setup="hmptree", message=FALSE, warning=FALSE, fig.width=10, fig.height=8}
f1 <- f + 
  new_scale_fill() +
  geom_fruit(
    data = df_ring_heatmap,
    geom = geom_tile,
    mapping = aes(x = Sites, y = ID, fill = Sites, alpha = Abundance),
    color = "grey50",
    size = 0.02,
    offset = 0.04 # control the spacing between the tree layer and heatmap layer
    ) +
  scale_fill_manual(values=c("#0000FF","#FFA500","#FF0000",
                             "#800000", "#006400","#800080","#696969")
                    ) +
  scale_alpha_continuous(range = c(0, 1)) +
  theme(
    legend.key.height = unit(.5, 'cm'), # legend 大小 
    legend.key.width = unit(.5, 'cm'),
    legend.title = element_text(size = 8),
    legend.text = element_text(size=6)
    ) 

f1
```

### 3.2. Aligning graph to the tree based on tree structure

```{r hmptree-barplot, exercise=TRUE, exercise.setup="hmptree", message=FALSE, warning=FALSE, fig.width=4, fig.height=8}
p.barplot <- ggplot() + 
  geom_col(
    data = df_barplot_attr, 
    mapping = aes(x=HigherAbundance, y = ID, fill=Sites)
    ) + 
  scale_fill_manual(values=c("#0000FF","#FFA500","#FF0000", 
                             "#800000", "#006400","#800080","#696969")
                    ) + 
  theme_tree2()

p.barplot
```

```{r hmptree-barplot2, exercise=TRUE, exercise.setup="hmptree-heatmap2", fig.width=10, fig.height=8, message=FALSE, warning=FALSE}
f2 <- f1 + 
  geom_fruit(
    data = df_barplot_attr,
    geom = geom_col,
    mapping = aes(x=HigherAbundance, y = ID, fill=Sites),
    pwidth = .38, # the space ratio for plotting the barchart relative to the tree layer
    axis.params = list(axis = 'x')
    ) +
  scale_fill_manual(values=c("#0000FF","#FFA500","#FF0000",
                             "#800000", "#006400","#800080","#696969")) 

f2
```

### 4. Transforming circular layout to rectangular layout

```{r hmptree-rect, exercise=TRUE, exercise.setup="hmptree-barplot2", fig.width=10, fig.height=8, message=FALSE, warning=FALSE}
f2 + layout_rectangular()
```

## Visuzalization of potential reassortment

### 1. Read in the two trees (H5 and N1 tree)


```{r h5n1, exercise=TRUE, message=FALSE, warning=FALSE}
library(treeio)
library(ggplot2)
library(ggtree)

H5 <- read.newick(system.file("extdata/csc/H5_mcc.nwk", package="learnggtree"))
N1 <- read.newick(system.file("extdata/csc/N1_mcc.nwk", package="learnggtree"))

# extract data used in ggtree plot
dH5 <- ggtree(H5)$data
dN1 <- ggtree(N1)$data

dH5
```

### 2. Flipping the N1 tree and merging the tip info of the two trees

```{r h5n1-2, exercise=TRUE, exercise.setup="h5n1", message=FALSE, warning=FALSE}
library(dplyr)

dN1$x <- max(dN1$x) - dN1$x + max(dH5$x) + 1

dd <- bind_rows(dH5, dN1) %>% filter(!is.na(label))
dd
```

### 3. Visualizing the two trees face-to-face

```{r h5n1-3, exercise=TRUE, exercise.setup="h5n1-2", fig.width=8, fig.height=6, message=FALSE, warning=FALSE}
pp <- ggtree(H5) + geom_tree(data=dN1)
pp
```

### 4. Connecting the taxa by lines

```{r h5n1-4, exercise=TRUE, exercise.setup="h5n1-3", fig.width=14, fig.height=8,out.width='95%', message=FALSE, warning=FALSE}
Cluster <- read.table(file = system.file("extdata/csc/taxa.cluster", package="learnggtree"), 
                      sep = '\t', header = TRUE)
Cluster_data <- merge(x=dd, y=Cluster, by.x="label", by.y = "name", all= TRUE)
Cluster_data

p3 <- pp + geom_line(aes(x, y, group=label, color=cluster), data=Cluster_data, alpha=.6)
p4 <- pp + geom_line(aes(x, y, group=label, color=geography), data=Cluster_data, alpha=.6)

aplot::plot_list(p3, p4, tag_levels = 'A')
```


## Phylogeographic example

see also <https://beast.community/workshop_continuous_diffusion_wnv#run-spread3>

### Load the required packages

```{r wnv, exercise=TRUE, message=FALSE, warning=FALSE}
library(treeio)
library(ggtree)
library(OpenStreetMap)
library(ggplot2)
library(paletteer)
```

### Draw the map

```{r wnv-1, exercise=TRUE, exercise.setup="wnv", fig.width=9.5, fig.height=4.5,out.width='95%', message=FALSE, warning=FALSE}
mp <- openmap(c(46,-125),
                c(21,-72),5,'apple-iphoto')

map_latlon <- openproj(mp, projection = "+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs")

pmap <- autoplot(map_latlon) 

pmap
```

### Display the tree structure on the tree


```{r wnv-2, exercise=TRUE, exercise.setup="wnv-1", fig.width=9.5, fig.height=4.5,out.width='95%', message=FALSE, warning=FALSE}

## read.tree
wnv_tree <- read.beast(system.file("extdata/csc/WNV_NA.tree", package="learnggtree"))

## visualize the tree on the map
pmap + 
    geom_tree(aes(x=location2, y=location1, alpha = height, color = height, size = height), 
          layout='slanted', data=wnv_tree, 
          arrow = arrow(length=unit(0.1, "inches")), 
          lineend = "round", linejoin = "round") +
   scale_alpha_continuous(range = c(0.5, 1),trans = "reverse") +
   scale_size_continuous(range = c(1.5, 0.3)) +
   guides(alpha = "none",
          size = "none") +
   scale_colour_paletteer_c("pals::ocean.thermal",  breaks = seq(0, 10, 10/7),
                    labels = c("2006-5", "2005-5", "2004-5", "2003-5", "2002-5", "2001-5","2000-5","1999-5"),
                    name = "Date") +
   theme(legend.key.height = unit(1, "null"),
        legend.justification.top = "right") +
   labs(x = "Longitude",
        y = "Latitude")
```

